The Noisy Work of Uncertainty Visualisation Research: A Review

### 2024-06-20 paper.qmd
a large amount of restructuring

<!-- TODO 
- convert to American spelling
- May not need to reference the same paper again and again
- add visually integrable to fig-tax once I work out what it is
- look through for papers that can be used to cite motivation for uncertainty visualisation experiments
-->

@@

There are several reasons provided to explain the difficulty in generalising the uncertainty visualisation literature. Some suggested or agree with the idea that visualisation typologies should move away from data types, uncertainty categories, and representation types and towards "task-centred typologies" [@Kinkeldey2014; @Hullman2016]. Other papers indirectly hint at this by arguing that the choice of best visualisation is highly dependent on the specific goal [@Griethe2006; @Spiegelhalter2017]. Others contend that visualisations are highly dependent on the audience and there is no such thing as a "best" visualisation that will be accessible to all members [@Kinkeldey2014; @Spiegelhalter2017]. These concerns appear across all application areas, and an underlying thread is that the problem arises from a lack of a cohesive and encompassing **definition of uncertainty**.

The review provided here attempts to address these issues. We highlight the three key issues facing the uncertainty visualisation literature, specifically the lack of a definition of uncertainty, the confusion as to what *is* an "uncertainty visualisation", and the difficulties with measuring "uncertainty". When disucssing each of these issues, we will discuss how this concept is understood in existing literature, why this understanding may be flawed, and the issues that arrise due to these flaws. By considering the *purpose* of visualizing uncertainty we avoid providing a simple list of the current methods, and instead consider if the current visualisation tools are sufficient to achieve that purpose and if not, where there is need for new methodology. The ideal role of this review to help synthesize this noisy field, by collecting together essential components, summarizing different viewpoints the current literature, and provide a guide for potential research topics.

# The purpose of uncertainty visualisation
## Other motivations 
## Signal supression

# When is signal supression appropriate?
Signal suppression is not always possible in a visualisation. Whether or not uncertainty should be used for signal-suppression largely depends on the purpose of the visualisation and the definition of uncertainty.

## What is uncertainty
### Current definitions of uncertainty
Most definitions of uncertainty treat it as a feature of the data, as something that exists in our data no matter it's purpose. This is not entirely accurate.

This view of uncertainty leads to a wide range of definitions that are designed to be encompassing, but struggle to capture the full picture. Some works [@Hullman2018; @Maceachren2012; @Thomson2005] focus narrowly on specific terms defined mathematically, such as probability, variance, error, or precision. Others [@Griethe2006; @Pang1997; @Pham2009; @Boukhelifa2017] include broader loosely related elements, such as missing values. Some authors use taxonomies to split uncertainty based on an endless stream of ever changing boundaries, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017; @Hullman2016; @utypo], if the uncertainty is in the  attribute, spatial elements, or temporal element of the data [@Kinkeldey2014], whether the uncertainty is scientific (e.g. error) or human (e.g. disagreement among parties) [@Benjamin2018], if the uncertainty is random or systematic [@Sanyal2009], statistical or bounded [@Gschwandtnei2016; @Olston2002], recorded as accuracy or precision [@Griethe2006; @Benjamin2018], which stage of the data analysis pipeline the uncertainty comes from [@utypo], how quantifiable the uncertainty is [@Spiegelhalter2017; @utypo], etc. In their own way, each of these approaches show an aspect of uncertainty that an author felt was important to differentiate.

The constantly shifting definitions make it hard to pin down and quantify uncertainty. The ramifications of this is seen throughout the literature, as a survey of visualisation authors cited "not knowing how to calculate uncertainty" as one of the primary reasons they did not include it in visualisations [@Hullman2020a].

The constanly shifting definitions might be a result of the view of uncertainty as a latent variable that can exist without an explicit purpose. This is not true, as uncertainty only exists as a by-product of inference.

### The case for uncertainty as a by product of inference
It is easy to see uncertainties relationship to inference when we consider what might *not* be considered uncertain rather than just trying to think about what *is*.

@@

## EDA vs Communication
## If signal supression is not the goal

# How should be test uncertainty visualisations?
## Uncertainty as signal
## Uncertainty as noise
## Other uncertainty experiments 
-----------------------------------------------

# UP TO HERE FOR SORTING

# What is uncertainty?

## Defining uncertainty with respect to inference

For example, @Boukhelifa2012 tried to quantify the strength of the intuitive connection between a line attribute called "sketchiness" and uncertainty. Participants were shown the six scenarios depicted in @fig-sketchy and asked to interpret what they believed the squiggly line indicates. The authors were aware that uncertainty seemed to have some "task" dependence but interpreted this as *context* dependence rather than dependence on a particular inferential statistic. Therefore, "sketchiness" was added to aspects of the visualization that had no obvious inferential statistic associated with it, so participants simply ignored the "sketchiness" or assumed it represented something else, such as an alternative option. In another experiment,

@Hofman2020 asked participants to judge the effectiveness of a particular treatment. One group was shown the prediction interval around the mean, which indicates the uncertainty associated with treatment *for a particular person* and another group was shown the the sampling interval which depicted the uncertainty associated with the *average effectiveness of the treatment*. There was no critical assessment by the authors on *why* these two distributions should be compared. The contribution of this experiment hinges on the belief that sharing a common descriptive statistic (in this case, the mean) is grounds to compare two unrelated inferential statistics, a belief which is clearly false when viewing uncertainty through the lens of inference.

![The graphics displayed by @Boukhelifa2012 to identify if there is an intuitive connection between sketchiness and uncertainty. These graphic were made without concern as to what (i.e. which statistic) in the image is supposed to be uncertain. This leads to the images being difficult to connect to uncertainty even if we assume tha that is what it is supposed to represent. For example, (S5) implies that participants should have read the plot to mean that the creators of a rail network map would be "uncertain" about the existence of a train line, something that is defies common logic.](sketchiness.png){#fig-sketchy width=50%}

Additionally, if we view uncertainty through the lens of inference then it is obvious that bias and variance both contribute to uncertainty regardless as to which stage of the analysis they appear. Many papers in uncertainty visualisation are not working from this null hypothesis. For example, @Padilla2021 found that high uncertainty in the model estimates (calculated uncertainty) and low forecaster confidence (which is typically an expression of suspected bias in the model) both caused participants to have decreased confidence in their results and suggested modelers express both if they are relevant. In another example, @Kale2019 discusses the importance of communicating decisions made in the data analysis pipeline and being aware of the alternatives. While there is nothing wrong with the results of these papers, the fact that they were published shows some level of surprise by these results. This work indicates that there was a significant number of authors in the field were not already aware that choices introduced early in the data analysis pipeline create bias and therefore uncertainty in our final values that should be communicated.

Existing mathematical definition of uncertainty (or related concepts) have been unable to unify this range of diverse concepts. For example, @Thomson2005 suggests a mathematical formula for *examples* of uncertainty, @Meng2014 mathematically defined the variance introduced to a model by the array of model choices, and information theory tries to quantify uncertainty using the idea of entropy. These existing methods are useful for calculating uncertainty as a specific stage of the analysis, but not thorough enough for an analyst to understand the causes of uncertainty and quantify it. The reality is that a lot of uncertainty visualisation authors do not seem to have an intuitive understanding of uncertainties connection to inference, when inference is being performed, and how to communicate this complicated by product of inference. This misunderstanding is not due to laziness or the fault of the authors, but rather is likely caused by the confusing set of existing definitions that leads authors to be unaware what they are supposed to be visualising.  Therefore, it is clear the field would greatly benefit from a strict definition of uncertainty. The task of providing an encapsulating mathematical definition of uncertainty is far beyond the scope of this work, however focusing on its relationship to inference is likely a fruitful step forward.

# What is an uncertainty visualisation?
Given that there is no definition of "uncertainty" readers may wonder how we can define "uncertainty visualisations". This is a reasonable issue given that you cannot quantify something that is not defined, and you can't visualise something you cannot quantify. This section discusses how the problems caused by the absent definition of uncertainty bleed into our understandings of uncertainty visualisation and creates a field that cannot clearly state what is or is not within its domain.

## Definitions of "uncertainty visualisation"
What exactly is or is not an uncertainty visualisation is just as poorly defined as the definition of uncertainty, so authors seem to decide for themselves what is an uncertainty visualisation. This appears in the literature as a mountain of seemingly conflicting statements about what is, or is not, an "uncertainty visualisation". For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, however at least one or both of these charts are used in a significant number of uncertainty visualisation experiments [@Ibrekk1987; @Olston2002; @Zhao2023; @Hofmann2012]. @Wickham2011 suggests their product plot framework, which includes histograms, should have a way to measure uncertainty, but does not consider that a histogram is *already* a depiction of PDF, something that is often considered an uncertainty visualisation by other authors. This completely untethered concept of an "uncertainty visualiation" leads to many user experiments where the visualisations used have a shocking information asymetry.

This collection of examples starts to paint a pretty clear picture. Visualisations with rather shocking information asymmetry are regularly compared because they are both "uncertainty visualisations". This results in a series of experiments where the visual aspects of the graphic are not even required to anticipate the experimental results. Ideally this problem would be avoided by defining the information in a graphic using *The Grammar of Graphics* since it is a sufficient summary of the information in a graphic and it allows a plot to be considered a statistic [@Vanderplas2020]. The process for summarising the information in a plot using *The Grammar of Graphics* is depicted in @fig-pipeline [@Leland2005]. Unfortunately, *The Grammar of Graphics* is not perfectly suited to summarise the information in an uncertainty visualisation since it assumes that we start from a finite data set. Very often uncertainty is expressed using a distribution or a resampling method that does not have an upper limit on the "sample size". Additionally many uncertainty visualisation experiments notice an interaction between sample size and graphic effectiveness [@Kale2018; @Newburger2022; @Hofmann2012], an effect that is likely caused by a difference in human perception of mass versus sample. Additionally, *The Grammar of Graphics* behaves strangely when the sample size is adjusted and it the distance between two graphics becomes less consistent. @fig-infintegraph shows how three graphics defined by *The Grammar of Graphics* can change as n approaches infinity. The two graphics that are very similar according to the *The Grammar of Graphics* diverge into different graphics, and two graphics that are very different, converge into the same graphic. This means the statistics visualised in the plot as well as the *The Grammar of Graphics* may not be an adequate to summarise the information in a plot that is being used for uncertainty visualisation. Some authors have noticed difficulties with visualising distributions using `ggplot2` and created extensions that make showing distributions more intuitive [@Pu2020; @Kay2023], however these extensions do not address the core discrepancies with uncertainty visualisation and *The Grammar of Graphics*. This does not mean *The Grammar of Graphics* should be abandoned when we attempt to summaise the information in an uncertainty visualisation, very often visations use the same sample to generate all their plots, so this issue is not always present, however these limitations should be kept in mind when analysing uncertainty visualisations using *The Grammar of Graphics*.

![(The grammar of graphics data analysis pipeline from @Leland2005 written in black, with the tidy data equivalent written in blue. The product plot framework and the grammar of probabilistic graphics frameworks are in green. It can be seen that the grammar of graphics provides a set of instructions that create a graphic, however the inferential power of this graphic is not considered and relies on many additional assumptions that are not defined by the grammar.](grammarofgraphicspipeline.jpeg){#fig-pipeline}

 ![This graphic shows three simple graphics made from the same data set that diverge according to the grammar of graphics. Despite being distinct graphics from a stage as low down as the statistic (or data manipulation in the tidy data framework), as our sample size increases to infinity, the grahics converge to the same distribution and become visually indistinguishable. ](infinitegraph.jpeg){#fig-infinitegraph}

An alternative way authors consider describing an uncertainty visualisation is to use taxonomies. However, these taxonomies tend to blur the line between the mathematical elements of uncertainty and the visual depiction of those mathematical objects. Some taxonomies of uncertainty highlight how confusing most authors find the concept, and how difficult it is to articulate the rules of an uncertainty visualisation when uncertainty itself if not defined. @Kinkeldey2014 categorised uncertainty according to five criteria depicted in @fig-tax which considers if a visualisation is implicit or explicit, intrinsic or extrinsic, visually integrable or seperable, coincident or adjacent, and static or dynamic. A similar version of this taxonomy was presented by @uncertchap2022 who commented that visualisation can be organised into two categories, "graphical annotations of distributional properties" and "visual encodings of uncertainty" which seems to functionally align with the intrinsic/extrinsic distinction by @Kinkeldey2014. @Griethe2006 organised uncertainty visualisations into two cases (1) a hypothesis test was preformed to confirm the validity of the visualisation and (2) the visualisation has uncertainty depicted. @Potter2012 organised several existing uncertainty visualisations into groups based on the dimensionality of the data (1D, 2D, 3D, and No Dimension) and the dimensionality of the uncertainty (Scalar, Vector, Tensor).  However, because the term "PDF", function that is used to describe a random variable, is used to describe both the data and the uncertainty for all dimensions. @Grewal2021 created a taxonomy that mapped uncertainty visualisations to some point in a 2D space defined by the "domain expertise" and "continuum of discreteness" (that scaled from "point estimate" to "continuous distribution").

@@

## Uncertainty visualisations should perform signal supression
It is rarely made clear exactly how uncertainty visualisation is different to normal information visualisation. Visualising a graphic is nothing more than computing a statistic, mapping that statistic to a visual feature and rendering that graphic. Nothing in this process suggests visualising an estimate of the variance should be any different to visualising an estimate of the mean, however this field *does* implicitly treat them different. Why?

Uncertainty visualisation papers usually provide one of two primary justifications for their existence. The first justification is that uncertainty is fundamentally differently to other variables due to the psychological heuristics involved in interpreting uncertainty. Therefore uncertainty visualisation is different to normal visualisation as authors must consider the psychological effects of what they are visualising [@Spiegelhalter2017; @Hullman2019]. The second justification assets that uncertainty is an additional variable and it is of vital importance to interpreting an estimate. Therefore uncertainty visualisation is a high dimensional visualisation problem, as we need to figure out how to seamlessly add uncertainty into already existing graphics [@moritz2017trust; @Griethe2006]. Why some authors pick one motivation over another is rarely explained and both leave the role of uncertainty in the visualisation unclear. The literature never explains if uncertainty should be treated as a variable, as metadata, or as something else entirely [@Kinkeldey2014]. @Hullman2016 commented that it is straightforward to show a value but it is much more complex to show uncertainty, but did not explain *why* Interestingly, these motivations do not remotely line up with the most common reason that people *outside* the uncertainty visualisation community thinks visualising uncertainty is important. The true reason that uncertainty visualisation is differerent to normal information visualisation, is that normal information visualisation is designed to *find* signals in our data, while uncertainty visualisation should be designed to *supress* it.

{{< pagebreak >}}

-----------------------------------------------

# UP TO HERE FOR RESTRUCTURE/REWRITE

### Interpretation and semantics
Interpretation and semantics experiments are seeking to identify a dimension (or visual task) that uncertainty naturally maps to. These experiments inherently view uncertainty as a variable that is separate to the variable on which we have mapped our signal. For example, lets say we have a map were maximum daily temperature is presented using points where the colour (red for hot and blue for cold) of the point is associated with the temperature, and the blurriness of that point is associated with the variance *of* that temperature. It is highly likely that our brains will not flatten that into a single variable depicting noise and signal, but rather *separately* extract the temperature (colour) and uncertainty (blur) information as two independent variables. If the variables are extracted separately, there is no guarantee that the uncertainty will act as an appropriate signal suppressor. This problem has been noticed by others in the field, that typically use this method (specifically in the spatial uncertainty context) and a desire for representations that integrate uncertainty and signal is one of the reasons for the invention of the value-suppressing uncertainty pallet [@Correll2018].

@@

# How do we measure uncertainty?
If uncertainty is inherently different, we would expect evaluation studies of uncertainty visualisations to produce different results to similar evaluation studies on "normal" variables. This does not seem to be the case.
Uncertainty is noise, but most evaluation experiments measure it as signal.
The previous section highlights a unique and fascinating problem faced by uncertainty visualisation. If asking direct questions about uncertainty causes us to treat it as a signal, how do we evaluate uncertainty as *noise*?
 When we ask the viewer of a plot to look at data and extract a value, we are asking them to perform inference on that value. There will be noise associated with that answer and that is uncertainty. If we ask direct question about some uncertainty metric, we have turned the uncertainty into signal because that is what the participants are drawing inference on.

## Measuring uncertainty as signal
Uncertainty visualisation papers can be organised according to the *goal* of the experiment. Evaluation experiments are the standard rule for visualisations because the human brain is not as reliable as mathematical calculation. Therefore, user studies often aim to assess the limitations, biases, and heuristics of our mental calculator so that we can better understand the problems we may encounter when we plot our data. This is not to say any paper that suggests a visualisation without an evaluation experiment is completely lacking in justification, and there are many papers that suggest a novel visualisation without an evaluation study. Sometimes these papers are a preliminary step in finding a solution for common problems and intend to evaluate the visualisation in later work. The reasoning for this is obvious. There are often heuristics and biases that are not obvious to us when designing visualisation. Additionally, these heuristics and biases can change depending on the larger scope of the graphic and the population we are communicating with [@Spiegelhalter2017; @Kinkeldey2014].

The overarching goal of most uncertainty visualisation evaluation papers belong to one of: performance (how effectively a participant can extract information from a plot), interpretation and semantics (the ease with which a particular visualisation is associated with uncertainty), and quality of user experience (if the participants liked the plot or not)[@Hullman2019]. @Hullman2019 found that the majority of papers evaluate visualisations on performance (approximately 65% of the papers they surveyed) or interpretation and semantics (approximately 17%) and both of these evaluation goals will run into problems because of their conceptualisation of uncertainty.

### Repeating perceptual task experiments
Given the previous section, one might consider problems in uncertainty visualisation to be fixed if we could restrict evaluation experiments to cases where the information in each graphic is almost identical. Unfortunately this still often results in uninteresting or obvious results.

@@

These results show that uncertainty is not technically different to any other variable. When trying to anticipate the results of these studies, we can use the same principles of information equivalence and difficulty of relevant mental tasks to understand which plots will outperform others. This leads us to wonder... why is uncertainty visualisation a field at all?

### Performance
There are many metrics experimenters use to evaluate the performance of a visualisation. Each metric, such as can use accuracy, decision quality, confidence, trust and others each have their own issues which prevent them from accurately capturing the effects of uncertainty in a visualisation.

By far the most common metric used is accuracy, which is used by approximately 36% of evaluation studies [@Hullman2019]. In previous sections, we pointed out issues that arise when accuracy (of extracting probability or variance) is used as the primary metric in evaluating uncertainty visualisations, the work simply replicates known concepts in information visualisation. Subjective metrics, such as how much participants liked the aesthetic representation of a visualisation, or metrics with an unclear motivation or use, such as how memorable the information in a plot is or how much a user interacted with a plot, are ignored in this section. Below we will explain in detail why experiments that try to measure uncertainty using methods such as decision making, trust, or other types of questions, also fail to capture the effects of visualising uncertainty in a plot.

#### Decision making and risk aversion

![Visualisation of the process of transforming raw data to a visualisation insight. Summarising the data using common statistics removes information from our data. Once the data is in a format that represents the varibales of interest, organising that information in a visualisation makes it easier to extract insights from these statistics and convert them to insights. If information is removed at the statistics stage, it cannot be added back in at the visualisation stage, as it is meerly an efficient technique to organise and present information to enable a large range of insights. ](visprocess.jpeg){#fig-location width=70%}

Decision making tasks are often described by authors as a more realistic version how plots are used in practice. While we do not disagree with this statement, exactly *why* these tasks are different to value extraction tasks is never explained. While the task may be more similar to how plots are actually used, that does not necessarily mean it is a better *experimental* environment for evaluating plots. Let us consider this in more detail. A normal value extraction task involves:

1) correctly interpret the question
2) extract the specified value
3) report the value

On the other hand, a decision making task using an uncertainty visualisation involves:

1) Correctly interpret the question
2) Use some risk utility function to set a threshold for an "acceptable level of risk"
3) Extract the value for from the plot that will be compared to the threshold
4) Make a judgement based off that value

The key aspect that separates a decision making task from a value extraction task, for uncertainty visualisations, is the inclusion of a utility function. While this might seem like a small change, it actually completely warps the experiment in unexpected.

The first issue is that individuals will all have their own individual risk utility function, that is, how much they want to avoid or engage with risks or uncertainty. Therefore decision making experiments that have an additional layer of noise that value extraction experiments do not. We cannot be sure if participants answered differently from the ground truth because a visualisation was difficult to read *or* because the participants risk utility function did not align with the one set by the authors. Several authors have offered solutions to this issue, however the problem is deeper than any of them seem to realise.

@Hullman2016 suggested providing a utility framework for each experiment, to instruct the participants in how to account for the uncertainty information. This is a method that seems to have been adopted by @Fernandes2018 who describe the following scenario to participants who are trying to maximise the coins in their experiment:
> Subjects gain coins for every minute they are able to continue an activity that is valuable to them (e.g., watching TV at home) before going to the bus stop, and gain a bonus for arriving at their intended destination early. Subjects incur a coin penalty for time spent waiting at a stop for a bus to arrive.

Any payment scheme that is incorporated into an experiment will also implicitly set up a utility framework, as it is assumed participants will try to maximise their payments. In tasks such as this, the ground truth of each question is typically selected to be the value a ration agent would select and visualisations are evaluated on the basis of how far the participants responses are from that ground truth. The problem with this, is that all the complexity of the question is in *working out* the best response, and it has little to do with the visualisation itself. The visualisation aspect of these studies becomes a value extraction experiment.

@Cheong2016 tested multiple different visual representations of uncertainty for representing the likelihood of a house being burned down based on its location. Their payment scheme, which paid out $0.10 for a correct choice (i.e. staying when the house was not burned down or leaving when the house was burned down) and 0 for an incorrect choice (i.e.leaving when the house didn't burn down or staying when the house burned down), meant participants were incentivised to base their entire leave/stay decision on whether or not the likelihood of a fire at their house is above or below 50%. If the participants *did* correctly identify the optimum strategy and answer accordingly, the "decision making experiment" was actually *just a value extraction experiment*. The decision making aspect ceased to matter, all that mattered was identifying if the probability of a fire was greater than 50%. Interestingly, despite this very obvious and simple tactic to maximize payout from the experiment, it seems like many participants did not adopt it, instead acting much more warily as though they were considering whether or not they would *actually* evacuate in the event of a fire. This means it is likely that introducing a payment scheme or a utility function creates too much mental labour for the participants to answer "correctly", and they would rather stick with their own risk utility function that feels *natural* than go through the mental labour of adopting another. Setting up an experiment that requires participants to do laborious calculations or strenuous mental effort for a small marginal benefit is unlikely to result in successful participant responses because it is not in the "spirit" of visualisation. (*Cite Gap 16: Examples of value extraction decision making tasks*)

It is clear that audiences primarily interact with visualisations through their "System 1" brain [@daniel2017thinking]. This aligns with other authors comments that visualisation is primarily about "gists" [Spiegelhalter2017]. In this sense, asking participants a question that requires them to shift to their "System 2" is entirely contrary to *why* we use visualisation. The exact reasons we describe visualisation to be so powerful, that they are able to communicate complicated information fast and efficiently *is a product of the system 1 brain* [@daniel2017thinking], however this means displaying information using visualisations have the same weakness' as system 1 thinking, and visualisation authors need to be aware of this. Asking participants to answer questions that require complicated calculations does not reflect how visualisations are used, and will likely be ignored by the participants. While we agree with @Hullman2016 who suggested that what determines an appropriate ground truth is largely a philosophical exercise, we differ in the sense that we do not think it should be done, lest you create a needlessly complicated value extraction task.

Additionally, several papers mention "risk-aversion" in participant decisions as a negative by-product of a particular visualisation design, however that displays a failure to understand that risk aversion is *not* a mistake, but rather it *reveals the utility of decreased uncertainty*. It may be interesting to investigate why a particular visualisation elicits increased risk-aversion, for example is it due to poorer estimates or increased awareness of negative outcomes, however simply reporting this as "risk aversion" conflates these factors. The work that advocates for transparency in risk-communication gives you whiplash when it's subtext argues that transparency leads to "incorrect" conclusions. There have been other discussions on the appropriateness of discussing risk-aversion as a bias [@Vranas2000], but few have made the point that if uncertainty holds some *value* there is not technically a "right" decision at all.

>>>>

<!-- TODO 
- convert to American spelling
- May not need to reference the same paper again and again
- Update background to reflect new structure of paper
-->

@@

There are several reasons provided to explain the difficulty in generalising the uncertainty visualisation literature. Some suggested or agree with the idea that visualisation typologies should move away from data types, uncertainty categories, and representation types and towards "task-centred typologies" [@Kinkeldey2014; @Hullman2016]. Other papers indirectly hint at this by arguing that the choice of best visualisation is highly dependent on the specific goal [@Griethe2006; @Spiegelhalter2017]. Others contend that visualisations are highly dependent on the audience and there is no such thing as a "best" visualisation that will be accessible to all members [@Kinkeldey2014; @Spiegelhalter2017]. These concerns appear across all application areas, and an underlying thread is that the literature on uncertainty visualisations have become disconnected from their purpose.

The review provided here attempts to address these issues. We start by discussing why uncertainty visualisation has become it's own sub-field, distinct from "normal" information visualisation. In this discussion we establish the purpose of uncertainty visualisation is to perform "signal-suppression", that is, hide statistically insignificant signals while promoting statistically rigorous signals. **EDIT FROM HERE** We then move highlight the three key issues facing the uncertainty visualisation literature, specifically the lack of a definition of uncertainty, the confusion as to what *is* an "uncertainty visualisation", and the difficulties with measuring "uncertainty". When disucssing each of these issues, we will discuss how this concept is understood in existing literature, why this understanding may be flawed, and the issues that arrise due to these flaws. By considering the *purpose* of visualizing uncertainty we avoid providing a simple list of the current methods, and instead consider if the current visualisation tools are sufficient to achieve that purpose and if not, where there is need for new methodology. The ideal role of this review to help synthesize this noisy field, by collecting together essential components, summarizing different viewpoints the current literature, and provide a guide for potential research topics. **TO HERE**

# The purpose of uncertainty visualisation
It is rarely made clear exactly how uncertainty visualisation is different to normal information visualisation. @Hullman2016 commented that it is straightforward to show a value but it is much more complex to show uncertainty, but did not explain *why*. A visualisation is nothing more than computing a statistic, mapping that statistic to a visual feature and rendering that graphic. Nothing in this process suggests visualising an estimate of the variance should be any different to visualising an estimate of the mean, however this field *does* implicitly treat them different. Why?

Uncertainty visualisation papers usually provide one of two primary justifications for their existence. The first justification is that uncertainty is fundamentally differently to other variables due to the psychological heuristics involved in interpreting uncertainty. Therefore uncertainty visualisation is different to normal visualisation as authors must consider the psychological effects of what they are visualising [@Spiegelhalter2017; @Hullman2019]. The second justification assets that uncertainty is an additional variable and it is of vital importance to interpreting an estimate. Therefore uncertainty visualisation is a high dimensional visualisation problem, as we need to figure out how to seamlessly add uncertainty into already existing graphics [@moritz2017trust; @Griethe2006].

Why some authors pick one motivation over another is rarely explained and both leave the role of uncertainty in the visualisation unclear. The literature never explains if uncertainty should be treated as a variable, as metadata, or as something else entirely [@Kinkeldey2014].

Trust is a by product of displaying uncertainty rather than the goal of it, and viewing the relationship in the converse direction can lead to misguided research. Considering trust, and not transparency, as the metric of importance in uncertainty communication can lead to a questionable subtext that argues against transparency, something that has been noticed by several other authors [@Spiegelhalter2017; ONeill2018]. @Hullman2020a found that author simultaneously argued that failing to visualise uncertainty was akin to fraud, but also many avoided uncertainty visualisation because they didn't want their work to come across as "untrustworthy". These authors are optimising *trust* rather than *transparency*, which means they opt to leave out uncertainty information when it does exactly what it is supposed to, decrease certainty in conclusions.

# Signal-Supression
The true reason that uncertainty visualisation is differerent to normal information visualisation, is that normal information visualisation is designed to *find* signals in our data, while uncertainty visualisation should be designed to *supress* it.

## Uncertainty does not exist without signal
It is obvious that we cannot design a visualisation for signal suppression without a clear signal. This may be considered a limitation by some readers, however it is less of a restriction when we recognize that *uncertainty* itself does not exist without a signal.

#### The current definitions of uncertainty
Uncertainty is a concept that is famously hard to pin down, with constantly shifting definitions that mean it is quantified differently from author to author.

The ramifications of this is seen throughout the literature, as a survey of visualisation authors cited "not knowing how to calculate uncertainty" as one of the primary reasons they did not include it in visualisations [@Hullman2020a]. Some works [@Hullman2018; @Maceachren2012; @Thomson2005] get around this definition issue by focusing narrowly on specific terms defined mathematically, such as probability, variance, error, or precision. Others [@Griethe2006; @Pang1997; @Pham2009; @Boukhelifa2017] include broader loosely related elements, such as missing values, but ignore these concepts when translating that uncertainty to a visualisation. Existing mathematical definition of uncertainty (or related concepts) have been unable to unify this range of diverse concepts. For example, @Thomson2005 suggests a mathematical formula for *examples* of uncertainty, and information theory tries to quantify uncertainty using the idea of entropy, but the disconnect between the broad concept of uncertainty and what we can reliably quantify remains a problem.

Some authors prefer to define uncertainty in qualitative terms, often through the use of taxonomies. In these cases, uncertainty is split using an endless stream of ever changing boundaries, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017; @Hullman2016; @utypo], if the uncertainty is in the attribute, spatial elements, or temporal element of the data [@Kinkeldey2014], whether the uncertainty is scientific (e.g. error) or human (e.g. disagreement among parties) [@Benjamin2018], if the uncertainty is random or systematic [@Sanyal2009], statistical or bounded [@Gschwandtnei2016; @Olston2002], recorded as accuracy or precision [@Griethe2006; @Benjamin2018], which stage of the data analysis pipeline the uncertainty comes from [@utypo], how quantifiable the uncertainty is [@Spiegelhalter2017; @utypo], etc. In their own way, each of these approaches show an aspect of uncertainty that an author felt was important to differentiate.

What is interesting about all these views of uncertainty, is that they all imply that it is a latent feature of our data. That is, it is something that can be revealed through exploration, and not something that needs to be explicitly calculated. The constant stream of new definitions that are unable to capture the true essence of uncertainty might be a result of this false assumption that underpins ever definition. Uncertainty cannot be defined without a signal because uncertainty does not exist without a signal, is a by-product of inference.

#### Uncertainty as a by-product of inference
It is easy to see uncertainties relationship to inference when we consider what might *not* be considered uncertain rather than just trying to think about what *is*.

@@

## EDA vs Communication
The concept of uncertainty as a by-product of inference, not a latent feature of data, extends naturally to visualisation.

### No Such thing as an "uncertainty visualisation"
If uncertainty is a by-product of inference, a visualisation being an "uncertainty visualisation" does not exist.

What exactly is or is not an uncertainty visualisation is just as poorly defined as the definition of uncertainty, so authors seem to decide for themselves what is an uncertainty visualisation. For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, however at least one or both of these charts are used in a significant number of uncertainty visualisation experiments [@Ibrekk1987; @Olston2002; @Zhao2023; @Hofmann2012]. @Wickham2011 suggests their product plot framework, which includes histograms, should have a way to measure uncertainty, but does not consider that a histogram is *already* a depiction of PDF, something that is often considered an uncertainty visualisation by other authors.

@Kinkeldey2014 categorised uncertainty according to five criteria depicted in @fig-tax which considers if a visualisation is implicit or explicit, intrinsic or extrinsic, visually integrable or seperable, coincident or adjacent, and static or dynamic. A similar version of this taxonomy was presented by @uncertchap2022 who commented that visualisation can be organised into two categories, "graphical annotations of distributional properties" and "visual encodings of uncertainty" which seems to functionally align with the intrinsic/extrinsic distinction by @Kinkeldey2014. @Griethe2006 organised uncertainty visualisations into two cases (1) a hypothesis test was preformed to confirm the validity of the visualisation and (2) the visualisation has uncertainty depicted. @Potter2012 organised several existing uncertainty visualisations into groups based on the dimensionality of the data (1D, 2D, 3D, and No Dimension) and the dimensionality of the uncertainty (Scalar, Vector, Tensor).  However, because the term "PDF", function that is used to describe a random variable, is used to describe both the data and the uncertainty for all dimensions. @Grewal2021 created a taxonomy that mapped uncertainty visualisations to some point in a 2D space defined by the "domain expertise" and "continuum of discreteness" (that scaled from "point estimate" to "continuous distribution").

@@

## Designing Visualisation for Signal supression
Interpretation and semantics experiments are seeking to identify a dimension (or visual task) that uncertainty naturally maps to. These experiments inherently view uncertainty as a variable that is separate to the variable on which we have mapped our signal. For example, lets say we have a map were maximum daily temperature is presented using points where the colour (red for hot and blue for cold) of the point is associated with the temperature, and the blurriness of that point is associated with the variance *of* that temperature. It is highly likely that our brains will not flatten that into a single variable depicting noise and signal, but rather *separately* extract the temperature (colour) and uncertainty (blur) information as two independent variables. If the variables are extracted separately, there is no guarantee that the uncertainty will act as an appropriate signal suppressor. This problem has been noticed by others in the field, that typically use this method (specifically in the spatial uncertainty context) and a desire for representations that integrate uncertainty and signal is one of the reasons for the invention of the value-suppressing uncertainty pallet [@Correll2018].  

@@

# Visualising uncertainty as signal
If uncertainty is inherently different, we would expect evaluation studies of uncertainty visualisations to produce different results to similar evaluation studies on "normal" variables. This does not seem to be the case.
Uncertainty is noise, but most evaluation experiments measure it as signal.
The previous section highlights a unique and fascinating problem faced by uncertainty visualisation.

Even though uncertainty does not exist without a signal, that does not mean every visualisation that includes a concept related to uncertainty (such as a visualisation or error or variance) must be designed for signal supression. However, if a visualisation is not designed for signal supression, it should be treated the same as any other variable. These visualisation are not "special" in the sense that signal supression visualisations are.

@Hullman2019 found that the majority of papers evaluate visualisations on performance (approximately 65% of the papers they surveyed) or interpretation and semantics (approximately 17%) and both of these evaluation goals will run into problems because of their conceptualisation of uncertainty.

By far the most common metric used is accuracy, which is used by approximately 36% of evaluation studies [@Hullman2019].

This collection of examples starts to paint a pretty clear picture. Visualisations with rather shocking information asymmetry are regularly compared because they are both "uncertainty visualisations".

@@

These results show that uncertainty is not technically different to any other variable. When trying to anticipate the results of these studies, we can use the same principles of information equivalence and difficulty of relevant mental tasks to understand which plots will outperform others. This does not mean that visualising uncertainty as a signal is incorrect or bad, it just means that uncertainty as a signal is no different from any other information visualisation.

- decision making experiments are usually displaying the wrong signals

# How should we test uncertainty visualisations?
If asking direct questions about uncertainty causes us to treat it as a signal, how do we evaluate uncertainty as *noise*?
 When we ask the viewer of a plot to look at data and extract a value, we are asking them to perform inference on that value. There will be noise associated with that answer and that is uncertainty. If we ask direct question about some uncertainty metric, we have turned the uncertainty into signal because that is what the participants are drawing inference on.

## Current attempts to measure uncertainty
There is also a swath of studies that are aware a question that boils down to a value extraction experiment, or a question that should be answered by literally ignoring uncertainty information is not what we want when we consider uncertainty visualisations. These papers often try to ask a question that should utilise both the uncertainty and signal in the response, however this is rarely what actually occurs. This method typically results in in cryptic or confusing questions that create a large amount of noise on the interpretation side of the analysis [@Hullman2016].
